\documentclass[12pt,a4paper,openright,twoside]{book}
\usepackage[utf8]{inputenc}
\usepackage{disi-thesis}
\usepackage{code-lstlistings}
\usepackage{notes}
\usepackage{shortcuts}
\usepackage{acronym}
\usepackage{listings-rust}
\usepackage{svg}
\usepackage{epstopdf}
\epstopdfDeclareGraphicsRule{.svg}{pdf}{.pdf}{%
    inkscape -z -D --file=#1 --export-pdf=\OutputFile
}
\AppendGraphicsExtensions{.svg}

\school{\unibo}
\programme{MSc in Computer Science and Engineering}
\title{Design and development of a Rust-based execution platform for Aggregate Computing}
\author{Micelli Leonardo}
\date{\today}
\subject{Pervasive Computing}
\supervisor{Prof. Viroli Mirko}
\cosupervisor{Dott. Aguzzi Gianluca}
\session{IV}
\academicyear{2022-2023}

% Definition of acronyms
\acrodef{iot}[IoT]{Internet of Things}
\acrodef{vm}[VM]{Virtual Machine}
\acrodef{ac}[AC]{Aggregate Computing}
\acrodef{fc}[FC]{Field Calculus}
\acrodef{cps}[CPS]{Cyber-Physical Systems}
\acrodef{cas}[CAS]{Collective Adaptive Systems}
\acrodef{jvm}[JVM]{Java Virtual Machine}
\acrodef{dsl}[DSL]{Domain Specific Language}
\acrodef{ast}[AST]{Abstract Syntax Tree}
\acrodef{pc}[PC]{Pervasive Computing}
\acrodef{rfid}[RFID]{Radio-Frequency Identification Technology}
\acrodef{fp}[FP]{Functional Programming}
\acrodef{scafi}[ScaFi]{Scala Fields}
\acrodef{rufi}[RuFi]{Rust Fields}


\mainlinespacing{1.241} % line spacing in main matter, comment to default (1)

\begin{document}

\frontmatter\frontispiece

\begin{abstract}
    The rapid expansion of the Internet of Things has led to the proliferation of computational resources in the physical world, which are now embedded in everyday objects and environments.
    The Aggregate Computing paradigm has emerged as a promising approach to tackle the complexity of designing and coordinating these systems, by shifting the focus from individual devices to
    programming the global behavior of whole computational collectives. The current state-of-the-art implementation of this paradigm is \ac{scafi}, which targets the Java Virtual Machine.
    Concurrently, the \ac{rufi} project aims to exploit the Rust programming language's features to provide a more lightweight and memory-efficient implementation of Aggregate Computing that can be executed inside resource-constrained, ``thin'' devices.
    In this paper, we will present the design and development of a Rust-based platform that will enable the distributed execution of \ac{rufi}-based aggregate programs, taking a step towards the goal of bringing aggregate computing to thin devices.
\end{abstract}

\begin{dedication} % this is optional
    ‘‘The most profound technologies are those that disappear. They
    weave themselves into the fabric of everyday life until they are
    indistinguishable from it.’’
    \begin{flushright}
        (Mark Weiser, 1991)
    \end{flushright}
\end{dedication}

\begin{acknowledgements} % this is optional
    I would like to express my deepest gratitude to my supervisor, Prof. Mirko Viroli, for his guidance and support throughout the development of this thesis.
    I also could not have undertaken this project without the support of my supervisor, Doct. Gianluca Aguzzi and the valuable advice and time he dedicated to me.
    I am also grateful to Doct. Nicolas Farabegoli and professors Roberto Casadei and Danilo Pianini for their help and support.
    A special thanks go to my group of friends first and colleagues later, the `Angels`'': Angelo, Paolo, Davide, Angela, Filippo, Eddie and Francesco.
    I would also like to thank my dear friend and fellow rustacean and gymbro Luca, as well as all the people who have been close to me during my years in Cesena: Massimo, Marco, Schiaro, Riccardo, Andrea and
    all the amazing people I met during this wonderful adventure.
    I'm also grateful to my amazing group of friends at home, my ``Branco'': Dona, Cristian, Riki, Claudio, Francesco, Simone, Aldo and Raffaele, we all met in childhood and have never been apart since.
    Last but not least, I would like to thank from the bottom of my heart my mother Simona and all of my family for being my best supporters and for always being there for me.
\end{acknowledgements}

%----------------------------------------------------------------------------------------
\tableofcontents
\listoffigures     % (optional) comment if empty
\lstlistoflistings % (optional) comment if empty
%----------------------------------------------------------------------------------------

\mainmatter

%----------------------------------------------------------------------------------------
% CHAPTERS
%----------------------------------------------------------------------------------------
\input{sections/chapter01.tex}
\input{sections/chapter02.tex}
\input{sections/chapter03.tex}
\input{sections/chapter04.tex}
\input{sections/chapter05.tex}
\input{sections/chapter06.tex}
\input{sections/chapter07.tex}
%----------------------------------------------------------------------------------------
% BIBLIOGRAPHY
%----------------------------------------------------------------------------------------

\backmatter

\nocite{*} % comment this to only show the referenced entries from the .bib file

\bibliographystyle{alpha}
\bibliography{bibliography}

\end{document}