%! Author = leona
%! Date = 09/02/24
% !TeX root = ../thesis-main.tex

\chapter{Analysis and Requirements}
\label{chap:requirements}
In this chapter, we will present the goal of this thesis and describe how it will be achieved. From here, an organized structure of requirements will be presented.

\section{Thesis' Goal}
\label{sec:goal}
This thesis is within the same scope and ambit as the RustFields project, i.e. enabling the execution of aggregated programs on thin devices that would not support the JVM.
Specifically, this thesis aims to continue in this direction through the development of three objectives:

\begin{enumerate}
    \item \label{obj:1} test, validate and improve the design of the RuFi-core module;
    \item \label{obj:2} develop a new module, RuFi-distributed, that will enable the distributed execution of aggregate programs within a network of devices;
    \item \label{obj:3} develop and test an aggregate program that can be executed on a network of devices.
\end{enumerate}

\section{Requirements Breakdown Structure}
\label{sec:rbs}
From the thesis' goal and objectives, it is possible to devise a set of tasks that need to be accomplished. Each task can be further refined in sub-tasks that constitute the requirements for the main task completion.
Following this approach, a complete breakdown of the requirements emerges, as shown in this section.

\begin{enumerate}
    \item \textbf{RuFi-core}: this requirement category is related to the objective \ref{obj:1}: testing and improving the RuFi-core module.
    \begin{enumerate}
        \item \textbf{Validation}:
            \begin{enumerate}
                \item expand the "by_round" test suite;
                \item implement functions to assert the equivalence of two aggregate programs;
                \item expand the test suite with "by_equivalence" tests;
            \end{enumerate}
    \end{enumerate}
    \item \textbf{RuFi-distributed}
    \item \textbf{Aggregate Program}
\end{enumerate}
